---
title: 'Inference for Paired Data'
author: "Josh Jarvey"
date: "05/29/2020"
output: word_document
fontsize: 12pt
---

Create a Word docx from this R Markdown file for the following exercise.  Submit the R markdown file and resulting Word docx file.

## Exercise 1  

To reduce ankle injuries, restrictive appliances such as taping and spatting (applying tape over the shoe and sock) have been employed. As part of a study at UWL, subjects also completed a 5-point Likert-type scale survey regarding their perceptions of the movement of each ankle appliance during exercise.

Researchers would like to compare the central values for perceptions of the movement of taped ankles compared to spatted ankles using $\alpha = 0.10$ and to estimate the difference with 90% confidence.

### Part 1a

Load the data set AnkleMovement.rda from the DS705 package.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
library(DS705data)
data("AnkleMovement")
```

---

### Part 1b	

Create a new variable of the differences, with the perceptions of the spatted ankle (spat) subtracted from the perceptions of the taped ankle (tape).

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
  #creating a vector of the differences: tape - spat
movement_difference = AnkleMovement$tape - AnkleMovement$spat

```

---

### Part 1c  

Create a boxplot and histogram for the sample of differences. 

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1c -|-|-|-|-|-|-|-|-|-|-|-

```{r}
boxplot(movement_difference, main = "Distribution of Percived Ankle Movement difference of Tape vs. Spatting", ylab = "Movement Difference")

hist(movement_difference,main = "Distribution of Percived Ankle Movement difference of Tape vs. Spatting", xlab = "Movement Difference")

```

---

### Part 1d  

Comment on the suitability of this data for the paired t-test, the Wilcoxon signed rank test, and the sign test.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1d -|-|-|-|-|-|-|-|-|-|-|-

Although the sample shows an outlier, the sample does appear to be generally normal, and the whiskers of the boxplot are of equal size. Because of this, I lean more toward the t-test with this sample over the Wilcoxon Rank Sign test since the sample does not appear symmetric with the median being skewed. 

---

### Part 1e 

Because the choice of test is somewhat unclear, as happens often in real life, try all three tests to compare the central values for subject's perceptions of the movement of taped ankles compared to spatted ankles using $\alpha = 0.10$.

Do the t-test first: 

#### Step 1

Define the parameter in words in the context of the problem.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e.step1 -|-|-|-|-|-|-|-|-|-|-|-

mu_d: The mean difference in percieved ankle movement score between Tape vs. Spatting for UWL athletes.


#### Step 2

State the null and alternative hypotheses for the test using the symbol you defined previously.
 
#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e.step2 -|-|-|-|-|-|-|-|-|-|-|-

H_0: mu_d  = 0
H_a: mu_d != 0

#### Step 3

Use R to generate the output for the test you selected.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e.step3 -|-|-|-|-|-|-|-|-|-|-|-

```{r}
t.test(AnkleMovement$tape,AnkleMovement$spat, alternative = "two.sided", paired = TRUE)

```

#### Step 4

State a statistical conclusion at $\alpha=0.10$ and interpret it in the context of the problem.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e.step4 -|-|-|-|-|-|-|-|-|-|-|-

At a 90% level of significance, there is not enough evidence to suggest that there is a difference between the percieved movement score of taped ankles vs. spatted ankles for UWL athletes. 

#### Step 5

Write an interpretation in the context of the problem for the 90% CI for the population mean difference.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e.step5 -|-|-|-|-|-|-|-|-|-|-|-

At a 90% level of significance, the population mean difference of percieved ankle movement scores among tape and spatting lie within -0.31 to 1.02 units for UWL athletes, thus indicating no significant difference. 

#### Step 6

Perform the Wilcoxon Signed Rank Test.  

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e.step6 -|-|-|-|-|-|-|-|-|-|-|-

```{r}
wilcox.test(AnkleMovement$tape, AnkleMovement$spat, alternative = "two.sided", conf.level = 0.90, conf.int = TRUE, paired = TRUE)

```

At a 90% level of significance, the population mean difference of percieved ankle movement scores among tape and spatting lie within -0.50 to 2.00 units for UWL athletes, thus indicating no significant difference. 



#### Step 7

Perform the sign test.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e.step7 -|-|-|-|-|-|-|-|-|-|-|-

```{r}
library(signmedian.test)
signmedian.test(movement_difference, mu=0, alternative = "greater", conf.level = 0.90)
```

At a 90% level of significance, the population mean difference of percieved ankle movement scores among tape and spatting lie within -1.00 to 2.00 units for UWL athletes, thus indicating no significant difference. 

---

#### Step 8

Construct a bootstrap confidence interval at a 90% level of confidence for the mean difference in population mean perception of movement for taped and spatted ankles.  Use a bootstrap sample size of 5000.  Compare this interval with the results of the 90% *t*-interval.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e.step8 -|-|-|-|-|-|-|-|-|-|-|-

```{r}
bootMeanPaired <- function(x,i){
  return(mean(x[i]))
}

library(boot)
boot.object <- boot(movement_difference, bootMeanPaired, R = 5000)

boot.ci(boot.object, type = "bca", conf = 0.90)

```

At a 90% level of significance, the population mean difference of percieved ankle movement scores among tape and spatting lie within -0.21 to 0.79 units for UWL athletes, thus indicating no significant difference. 

---

#### Step 9

Compare the results of the three hypothesis tests and also whether or not the 90% bootstrap interval agrees with the result of each test.  Which procedure should be reported and why?

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e.step9 -|-|-|-|-|-|-|-|-|-|-|-

All three hypothesis tests agree with each other, as well as the bootstrapped version - there doesnt appear to be a significance in the mean population mean difference of percieved ankle movement scores among tape and spatting for UWL athletes. Because they all agree with each other, I will report the t.test() since this is the most widely recognized test. 


## Exercise 2

In a nationwide study of insurance claims (in dollars) filed in the previous year, a random sample of 125 claims was selected from all claims for vehicles classified as small, meaning the gross vehicle weight rating (GVWR) is less than 4500 pounds.  

For each claim, the insurance company's estimate for the claim was also provided.

The data frame SmallAuto.rda contains the claims and estimates for each vehicle class. 

### Part 2a  

Load the data SmallAuto from the DS705 package.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
data("SmallAuto")

```

---

### Part 2b

Construct histograms and boxplots for both the estimated claims and actual for the small class of vehicle. Describe the shapes of these distributions.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
estimated = SmallAuto[SmallAuto$Category == "Estimate",1]
actual = SmallAuto[SmallAuto$Category == "Actual",1]

boxplot(SmallAuto$Claim~SmallAuto$Category, main = "Distribution of Small Auto Claims", ylab = "Price in $", xlab = "Claim Type")

hist(estimated,main = "Distribution of 'Estimated' Insurance Claims", xlab = "Price in $")
hist(actual,main = "Distribution of 'Actual' Insurance Claims", xlab = "Price in $")
```

The data in both categories display a skew to the right, with the whisker on the boxplot being more heavily weighted to the right. The data also have a moderate amount of outliers. 


---

### Part 2c  

Create a new variable of the differences for small vehicles, with the difference being the estimated claim amount minus the actual claim amount.  The estimated claim amounts in the first half of the vector are paired with the actual claim amounts in the second half of the vector so that row 1 and row 126 form a pair, rows 2 and 127, etc. 

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2c -|-|-|-|-|-|-|-|-|-|-|-

```{r}
claim_difference = estimated - actual

```

---

### Part 2c  

Create a boxplot, histogram, and normal probability plot for the sample of differences. Also, obtain the P-value for a Shapiro-Wilk normality test.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2c -|-|-|-|-|-|-|-|-|-|-|-

```{r}
boxplot(claim_difference, main = "Distribution of Small Auto Claims", ylab = "Price in $", xlab = "Claim Type")

hist(claim_difference,main = "Distribution of 'Estimated' Insurance Claims", xlab = "Price in $")

qqnorm(claim_difference)
qqline(claim_difference)




```

---

### Part 2d  

Comment on the shape of the distribution of differences and the suitability of this data for the paired *t*-test, the Wilcoxon signed rank test, and the sign test.  Which test will you use?

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2d -|-|-|-|-|-|-|-|-|-|-|-

Replace this text with your answer here.


---

### Part 2e 
    
Conduct an appropriate test to see if the population central values for the estimated claim amount is less than for the actual claim amounts for vehicles in the small class using $\alpha = 0.05$.

#### Step 1

Define the parameter in words in the context of the problem.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 2e.step1 -|-|-|-|-|-|-|-|-|-|-|-

Replace this text with your answer here.

#### Step 2

State the null and alternative hypotheses for the test using the symbol you defined previously.
 
#### -|-|-|-|-|-|-|-|-|-|-|- Answer 2e.step2 -|-|-|-|-|-|-|-|-|-|-|-

Replace this text with your answer here.

#### Step 3

Use R to generate the output for the test you selected. Pay close attention to the order of subtraction done behind the scenes in R.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 2e.step3 -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

#### Step 4

State a statistical conclusion at $\alpha=0.05$ and interpret it in the context of the problem.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 2e.step4 -|-|-|-|-|-|-|-|-|-|-|-

Replace this text with your answer here.

---

### Part 2f  

Write an interpretation in the context of the problem for a 95% two-sided confidence interval.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2f -|-|-|-|-|-|-|-|-|-|-|-
```{r}
# Insert your R code here.

```

Replace this text with your answer here.

---

## Exercise 3

The data frame AutoIns is very similar to Small Auto.  

In a nationwide study of insurance claims filed in the previous year, a random sample of 125 claims was selected from all claims for vehicles classified as small, meaning the gross vehicle weight rating (GVWR) is less than 4500 pounds  A separate sample of 125 claims for vehicles classified as standard, meaning the GVWR is between 4500 and 8500 pounds.

For each claim, the insurance company's estimate for the claim was also provided.

The data frame AutoIns.rda contains the claims and estimates for each vehicle class. The variables in the data frame are defined as follows:

claim.small = the actual claim amount in dollars for a vehicle in the small class

est.small = the estimated claim amount in dollars for a vehicle in the small class

claim.standard = the actual claim amount in dollars for a vehicle in the standard class

est.standard = the estimated claim amount in dollars for a vehicle in the standard class

### Part 3a  

Load the data AutoIns from the DS705 package and look at the structure of the data in the file.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 3a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

---

### Part 3b  

Is the data "stacked" or "side-by-side" ("tall" or "wide")?

### -|-|-|-|-|-|-|-|-|-|-|- Answer 3b -|-|-|-|-|-|-|-|-|-|-|-

Replace this text with your answer here.

---

### Part 3c  

Which pairs of variables in the data frame are independent of each other?  You can use the variable names to identify them.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 3c -|-|-|-|-|-|-|-|-|-|-|-

Replace this text with your answer here.

---

### Part 3d  

Which pairs of variables in the data frame are paired (matched pairs)? You can use the variable names to identify them.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 3d -|-|-|-|-|-|-|-|-|-|-|-

Replace this text with your answer here.

---