
---
title: 'Simple and Multiple Linear Regression'
author: "Josh Jarvey"
date: "06/12/2020"
output: word_document
fontsize: 12pt
---

Knit a Word file from this R Markdown file for the following exercises.  Submit the R markdown file and resulting Word file.   

## Exercise 1

The data for this problem comes from a dataset presented in Mackowiak, P. A., Wasserman, S. S., and Levine, M. M.  (1992), "A Critical Appraisal of 98.6 Degrees F, the Upper Limit of the Normal Body Temperature, and Other Legacies of Carl Reinhold August Wunderlich," Journal of the American Medical Association, 268, 1578-1580.  Body temperature (in degrees Fahrenheit) and heart rate (in beats per minute) were two variables that were measured for a random sample of 130 adults.  A simple linear regression was used to see if body temperature had an effect on heart rate.

The data are in the file normtemp.rda in the DS705data package, this data is included in the DS705data package so you can access it by loading the package and typing data(normtemp).

### Part 1a

Create a scatterplot with heart rate in the vertical axis and plot the estimated linear regression line in the scatterplot. Include descriptive labels for the x and y-axes (not just the variable names as they are in the data file). 

Note:  this data set needs a little cleaning first.  The heart rates are missing for two of the rows.  You can delete these rows from the data frame using the R function na.omit().  Just put the name of the data frame in the parenthesis.

Does it appear that a linear model is at least possibly a plausible model for the relationship between hear rate and body temperature?  Explain your answer.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
library(DS705data)
data("normtemp")
normtemp = na.omit(normtemp)
plot(normtemp$temp,normtemp$hr, main = "Heart Rate as a function of Body Temp",xlab = "Body Temp", ylab = "Heart Rate")
```

From a visual perspective, there does appear to be a weak positive linear pattern between heart rate and body temp. It does seem pretty weak however. 

---

### Part 1b

Write the statistical model for estimating heart rate from body temperature, define each term in the model in the context of this application, and write the model assumptions. (Note: the statistical model is the underlying true, but unknown, model for the population that includes the error or noise term.  The model obtained in 1c, is our estimate, obtained using least-squares regression, of the the deterministic (non-random) part of the true model.)

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1b -|-|-|-|-|-|-|-|-|-|-|-
y_hat = beta0 + beta1 * x + epsilon

  y_hat: The predicted heart rate
  beta0: The intercept heart rate (when there is no body temp)
  beta1: The coefficient of body temp
      x: The body temp provided
epsilon: The unknown error


Assumptions to be made:
1. Errors have a mean of 0
2. Errors have the same variance for all the x's
3. Errors are independent from each other
4. Errors are normally distributed

---

### Part 1c  

Obtain the estimated slope and y-intercept for the estimated regression equation and write the equation in the form hr$=\hat{\beta_0} + \hat{\beta_1}temp$ (only with $\hat{\beta_0}$ and $\hat{\beta_1}$ replaced with the numerical estimates from your R output).

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1c -|-|-|-|-|-|-|-|-|-|-|-

```{r}
    #create a linear regression between heart rate vs. temp. 
model = lm(hr~temp, data = normtemp)
    #grab the coefficients.
model$coefficients
```

Replace the ## symbols with your slope and intercept.

$\widehat{\text{hr}}$ = -179.12 + 2.57temp  

---

### Part 1d

Test whether or not a positive linear relationship exists between heart rate and body temperature using a 5% level of significance.  State the null and alternative hypotheses, test statistic, the p-value, and conclusion.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1d -|-|-|-|-|-|-|-|-|-|-|-

```{r}
    #check summary of the model.
summary(model)
```

H0: There is no linear relationship between heart rate and body temperature (beta1 = 0)
Ha: There is a linear relationship between heart rate and body temperature (beta1 != 0)

t-statistic = 2.878
p-value = 0.0047

At a 95% level of significance, there is enough evidence to claim that there is a linear relationship between the heart rate for all adults and body temperature (p=0.0047).
---

### Part 1e

Provide a 95% confidence interval to estimate the slope of the regression equation and interpret the interval in the context of the application (do not us the word “slope” in your interpretation).

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e -|-|-|-|-|-|-|-|-|-|-|-

```{r}
    #generate a 95% confidence interval for the variables in my model.
confint(model)
```

At a 95% level of confidence, the population mean heart rate for all adults increases from 0.80 to 4.34 bpm for each additional degree in body temperature. 

---

### Part 1f

Provide a 95% confidence interval to estimate the mean heart rate for all adults with body temperature $98.6^o$ F.  Interpret the interval in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1f -|-|-|-|-|-|-|-|-|-|-|-

```{r}
    #using the model, predict a temp of 98.6 degrees and provide a confidence interval.
predict(model, data.frame(temp=98.6), interval = "confidence")

```

At a 95% level of confidence, the population mean heart rate for all adults with a body temperature of 98.6 degrees is between 73.30 and 76.08 bpm.

---

### Part 1g
    
Provide a 95% prediction interval to estimate the expected heart rate for a randomly selected adult with body temperature $98.6^o$ F.  Interpret the interval in the context of the problem.
 
### -|-|-|-|-|-|-|-|-|-|-|- Answer 1g -|-|-|-|-|-|-|-|-|-|-|-

```{r}
    #using the model, predict a temp of 98.6 degrees and provide a prediction interval.
predict(model, data.frame(temp=98.6), interval = "prediction")

```

At a 95% level of confidence, the predicted heart rate for all adults with a body temperature of 98.6 degrees is between 60.96 and 88.43 bpm.

---

### Part 1h

Obtain the residuals and plot them against the predicted values and also against the independent variable.  Also construct a histogram, normal probability plot, and boxplot of the residuals and perform a Shapiro-Wilk test for normality.  Based on your observation of the plot of residuals against the predicted values, does the regression line appear to be a good fit?  Do the model assumptions appear to be satisfied?  Comment. 

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1h -|-|-|-|-|-|-|-|-|-|-|-

```{r}
    #pull out the residuals from the model
resids = model$residuals
    #pull out the fitted values from the model
pred = model$fitted.values
    #pull out the independent variable, temp, from the dataset
temp = normtemp$temp

    #plot independent variable vs. residuals
plot(temp, resids, main = "Temp (independent) vs. Residuals", xlab = "Temperature", ylab = "Residuals")
    #plot fitted vs. residuals
plot(pred, resids, main = "Fitted vs. Residuals", xlab = "Fitted values", ylab = "Residuals")
    #plot a histogram of the residuals
hist(resids, main = "Distribution of Residuals", xlab = "Residuals", ylab = "Frequency")
    #boxplot of the residuals
boxplot(resids, main = "Distribution of Residuals", ylab = "Residuals")
    #QQ Plot of the residuals
qqnorm(resids, main = "QQ Plot of Residuals")
qqline(resids)
    #test for normality of the residuals
shapiro.test(resids)

```

The linear model does appear to be a good fit since the residuals are have equal variance given both the independent variable and the fitted values. Additionally, the residuals appear to be normally distributed by both inspecting visually (histogram, boxplot, QQ Plot) and with a Shapiro test (p=0.6027).

---

### Part 1i

Examine the original scatterplot and the residual plot. Do any observations appear to be influential or be high leverage points?  If so, describe them and what effect they appear to be having on the estimated regression equation.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1i -|-|-|-|-|-|-|-|-|-|-|-

Replace this text with your answer here

---

### Part 1j

Perform the F test to determine whether there is lack of fit in the linear regression function for predicting heart rate from body temperature.  Use $\alpha = 0.05$.  State the null and alternative hypotheses, test statistic, the p-value, and the conclusion.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1j -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your answer here.

---

### Part 1k

Conduct the Breusch-Pagan test for the constancy of error variance.  Use α = 0.05.  State the null and alternative hypotheses, test statistic, the P-value, and the conclusion.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1k -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your answer here.

---

### Part 1l

Calculate and interpret the Pearson correlation coefficient $r$.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1l -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your interpretation here.

---

### Part 1m

Construct a 95% confidence interval for the Pearson correlation coefficient $r$.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1m -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

---

### Part 1n

Calculate and interpret the coefficient of determination $r^2_{yx}$ (same as $R^2$).

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1n -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your interpretation here.

---

### Part 1o

Should the regression equation obtained for heart rate and temperature be used for making predictions?  Explain your answer.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1o -|-|-|-|-|-|-|-|-|-|-|-

Replace this text with your answer here.

---

### Part 1p

Calculate the Spearman correlation coefficient $r_s$ (just for practice).

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1p -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

---

### Part 1q

Create 95% prediction and confidence limits for the predicted mean heartrate for each temperature given in the sample data and plot them along with a scatterplot of the data. (Look for the slides titled "Confidence Bands" in the presentation.)

```{r}
# Insert your R code here.

```

---

## Exercise 2

A personnel officer in a governmental agency administered three newly developed aptitude tests to a random sample of 25 applicants for entry-level positions in the agency.  For the purpose of the study, all 25 applicants were accepted for positions irrespective of their test scores.  After a probationary period, each applicant was rated for proficiency on the job.  

The scores on the three tests (x1, x2, x3) and the job proficiency score (y) for the 25 employees are in the file JobProf.rda.

(Based on an exercise from Applied Linear Statistical Models, 5th ed. by Kutner, Nachtsheim, Neter, & Li)

### Part 2a

We'd like to explore using interaction terms in a statistical model 
including the three first-order terms and the three cross-product interaction terms:

$$y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_1 x_2 + \beta_5 x_1 x_3 + \beta_6 x_2 x_3 + \epsilon$$

Use R to find the corresponding estimated model and also obtain the `summary()`.

## -|-|-|-|-|-|-|-|-|-|-|- Answer 2a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

---

### Part 2b

Use R to compute the VIF for each term in the model.  Are any of the VIFs over 10?  (We need to add this into Lesson 6, but it's covered in the Lesson 8 Swirl - I've put an example in the chunk below.  Replace the chunk with code to find the VIF's for this model.)

## -|-|-|-|-|-|-|-|-|-|-|- Answer 2b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Change this code so it computes VIF's for the model in the problem
# if necessary install.packages('HH'), but do it in the console and reply 'n'
# if it asks if you want to compile a binary package
# in this case the low VIF's indicate collinearity is not a problem for the terms
require(HH)
require(DS705data)
data(HealthExam)
fit <- lm(Weight~Height+Waist+SysBP,HealthExam)
vif(fit)
```

Replace this text with your answer here.

--- 

### Part 2c

The model from 2a is suffering from the effects of collinearity (which you should see in 2b), which inflates the standard errors of the estimated coefficients.

Using the model summary from 2a what do you notice about the overall model p-value (from the F-statistic) and the individual p-values for each term in the model?  Does it make sense that the overall model shows statistical significance but no individual term does?  

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2c -|-|-|-|-|-|-|-|-|-|-|-

Replace this text with your answer here.

---

### Part 2d

Use R to estimate and `summary()` the first order model corresponding to 

$$y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \epsilon$$

Is the first order model significant?

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2d -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your answer here.

---

### Part 2e

Do the interaction terms in 2a really add anything significant beyond the first order model in 2d?  Now we'll compare the models with and without interaction terms to see if the interaction terms make a statistically significant improvement to the fit of our models.

Test the significance of all three coefficients for the interaction terms as a subset by using `anova()` to compare the model from Part 2a to the first order model from Part 2d. Use a 5% level of significance.  State $H_0$ and $H_a$ and provide the R output as well as a written conclusion which includes the P-value.  Should we keep the interaction terms?

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2e -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your answer here.

---

### Part 2f

There are more methodical approaches to exploring different models that we'll learn about in a later lesson, but we'll try one more model here to get a bit more experience.  In this case we'll add a quadratic term $x_2^2$.  To do this you'll want to create a new variable `x2sq = x2^2` and include it in your model.  Use R to estimate and `summarize()` the model corresponding to: 

$$y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_2^2 +\epsilon$$

Examine the p-value corresponding to the quadratic term.  If the quadratic term is significant at significance level $\alpha = 0.05$, then according to the hierarchical approach we should retain it and the $x_2$ term.  If it isn't significant, then we won't retain it but we'll have to evaluate the significance of the $x_2$ term separately.

Should the quadratic term be retained in the model at a 5% level of significance?  

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2f -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your answer here.

---

### Part 2g

If you've been successful so far, then you should realize that the none of interaction terms nor the quadratic term have been significant (if you concluded otherwise, then review your work). This brings us back to the first order model in Part 2d.  Look at that model summary again.  There should be one term that is insignificant so omit it and use R to estimate our final and smaller first order model.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2g -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

---

### Part 2h

From the final first order model in 2g, obtain a 90% confidence interval for the coefficient of $x_3$ and interpret it in the context of this problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2g -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your answer here.

---

### Part 2i

Using the final first order model from 2g, construct a 95% prediction interval for a randomly selected employee with aptitude scores of $x_1=99, x_2=112,$ and $x_3=105$ to forecast their proficiency rating at the end of the probationary period. Write an interpretation for the interval in the context of this problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2i -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your answer here.

---

## Exercise 3

A research professor in a leading department of education is studying three different methods of teaching English as a second language. After three months in the program the participants take an exam and let $y$ be the score on the exam. The following model was used to assess the efficiencies of the three methods

 $$y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon$$
where 

$$x_1 =\begin{cases} 1 & \text{if Method 2} \\
                     0 & \text{otherwise}
       \end{cases}$$
       
and

$$x_2 =\begin{cases} 1 & \text{if Method 3} \\
                     0 & \text{otherwise}
       \end{cases}$$




### Part 3a

Use data in the file English.rda in the DS705data package to estimate the coefficients for the model:

$$y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon$$ 

Obtain the estimated intercept and coefficients and state the estimated mean English proficiency scores for each of the 3 methods of teaching English as a second language.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 3a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your answer here.

---

### Part 3b  

The researcher has given each participant a test prior to the beginning of the study and obtained an index, $x4$, of the participant’s English proficiency.

$$y=\beta_0 + \beta_1 x_4 + \beta_2 x_1 + \beta_3 x_2 + \beta_5 x_1 x_4 + \beta_6 x_2 x_4 + \epsilon$$

Using the estimated coefficients, write three separate estimated models, one for each method, relating the scores after 3 months in the program ($y$) to the index score prior to starting the program ($x_4$).

### -|-|-|-|-|-|-|-|-|-|-|- Answer 3b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
# Insert your R code here.

```

Replace this text with your answer here.